Subsidiary Recursion in Coq

### 2022-02-03 paper/paper.tex
Additional polishing through Section 3.

This first example does not invoke subsidiary recursions, but will
itself be used as a subsidiary recursion in other examples to follow.
Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix.  This is the typing for a top-level recursion.  More
generally, though, given a type \verb|R : Set| along with a
fold function for that type (i.e., of type
\verb|FoldT (Alg (ListF A)) R|), we would like to map an input list of
type \verb|R| to a pair of type \verb|list A * R|.  The first
component of this pair is going to be built up from scratch, and so
cannot have type \verb|R|; we cannot statically ensure that outer
recursions on it are legal.  But the second component will be a
subdatum of the input list, and so can still have type \verb|R|,
enabling outer recursive calls.  So we want:
\begin{minted}{coq}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{minted}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{minted}{coq}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{minted}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|ListF A R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{minted}{coq}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{minted}
\noindent The idea is that the recursion will signal if it is in the
one tricky case where \verb|p| does not match the first element, by
returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This suffix is available to the algebra
in the subdata structure of type \verb|ListF A R|.

@@

Figure~\ref{fig:spanalg} show the algebra
\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
\verb|SpanF| of the required type \verb|Set -> Set|.  We use
\verb|rollAlg| to create an algebra from something whose type is an
application of \verb|AlgF|.  This takes in all the components of the
recursion universe: the abstract type \verb|R|, the fold function
(\verb|fo|) for any subsidiary recursions (not needed here), a
function we choose to name \verb|span| for making recursive calls, and
finally \verb|xs : ListF A R|.  The algebra pattern-matches on this
\verb|xs|.  In the cases where it is empty or where its head
(\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
This signals to the caller that we really wished to return
\verb|([],xs)|, but could not because we do not have \verb|xs| at type
\verb|R|.  If the head does satisfy \verb|p|, then we recurse on the
tail (\verb|tl : R|) by calling the provided
\verb|span : R -> SpanF R|.  If \verb|span tl| returns
\verb|SpanNoMatch|, that means that we should make \verb|tl| the
suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
have \verb|tl : R| here, so we can do this.  In either case (for
return value of \verb|span tl|), we add the head to the front of the
prefix.

@@

\subsection{The \texttt{wordsBy} function (\texttt{WordsBy.v})}

Let us now see how to write \verb|wordsBy|, our example function from
Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.  The code
is in Figure~\ref{fig:wordsby}, assuming a type \verb|A : Set|.  The
setup is similar to that for \verb|span|.  We first define an algebra
\verb|WordsBy|, parametrized by the predicate \verb|p|, of type
\verb|Alg (ListF A) (Const (list (list A)))|.  This says that
\verb|WordsBy p| is an algebra (\verb|Alg|) for the \verb|ListF A|
functor, with carrier \verb|Const (list (list A))|.  \verb|Const| is
a combinator for creating the object part of constant functors;
\verb|FunConst| creates the morphism part (i.e., the \verb|fmap|
function).  We use \verb|Const| where the return type of the algebra
will not depend on its abstract type \verb|R|.  Here, we are
constructing from scratch a list of lists, so it will not be legal to
recurse on the list itself, or its (list) elements.  So we just use
the \verb|list| type of Coq's standard library.

The code for \verb|WordsBy| is essentially
the same as what we saw in Section~\ref{sec:intro}.  We pattern match
on \verb|xs : ListF A R|.  Recall that for this function, we are
trying to drop elements which satisfy \verb|p|, and return a list of
the sublists between maximal sequences of such elements.  In the
\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we are
supposed to drop it and recurse.  This is legal, because \verb|tl : R|
and \verb|wordsBy : R -> list (list A)|.  In the \verb|else| case,
we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
does not satisfy \verb|p|, and the remaining suffix \verb|z|.

Here we see the benefit of our approach.  From Figure~\ref{fig:span},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
comes from the type \verb|FoldT (ListF A) Alg R| of
\verb|fo|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
can indeed apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.  The figure also shows the code for the subsidiary
recursion \verb|wordsByr|.

\begin{figure}
\begin{minted}{coq}
  Definition WordsBy(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs =>
      match xs with
        Nil => []
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p) xs.
\end{minted}
\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
  fold (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

In this example, we see how to write a combinator that factors out a
subsidiary recursion.  The Haskell library \verb|Data.List.Extra| has
a function \verb|repeatedly|, defined essentially as in
Figure~\ref{fig:mapthroughhs}, though we attempt a more informative
name.  This is like the standard \verb|map| function on lists, except
that the function \verb|f| that we are mapping (or ``mapping
through'') takes in not just the current element \verb|a|, but also
the tail \verb|as|.  It then returns the value \verb|b| to include in
the output list, and whatever other list it wishes, upon which
\verb|mapThrough| will recurse.

To write this combinator using our infrastructure for subsidiary
recursion, we need to supply the mapped function with the fold
function for \verb|mapThrough|'s recursion.  This is so that the
mapped function can initiate a subsidiary recursion, returning a value
in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  So
the type we will use for mapped functions is:
\begin{minted}{coq}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{minted}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input.

Given this definition, the code is in Figure~\ref{fig:mapthrough}.
\verb|MapThroughAlg| is similar to the
Haskell code above, though when we call \verb|f|, we must
supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
from the definition of \verb|mappedT|, we have that \verb|b : B| and
\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> list B|
on \verb|c|.  Note that as we are building up a new list from scratch
(rather than just extracting some tail of the input list), we just
return \verb|list B|; we cannot perform further subsidiary recursion
on the output.

\begin{figure}
\begin{minted}{coq}
Definition MapThroughAlg{B : Set}(f:mappedT A B)
  : Alg (ListF A) (Const (list B)) :=
  rollAlg (fun R fo mapThrough xs =>
    match xs with
      Nil => []
    | Cons hd tl =>
      let (b,c) := f R fo hd tl in
        b :: mapThrough c
    end).
Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                      {B : Set}(f:mappedT A B) : R -> list B.
Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B.
\end{minted}
\caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
  code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
  we omit it (\texttt{MapThrough.v})}
\label{fig:mapthrough}
\end{figure}

@@

Finally, we have an example using our \verb|mapThrough| combinator
together with a subsidiary recursion, to implement \emph{run-length
encoding}.  This is a basic data-compression algorithm where maximal
sequences of $n$ occurrences of element $e$ are summarized by the pair
$(n,e)$~\cite{datacomp}.  Haskell code is in Figure~\ref{fig:rlehs}.
Recall that \verb|(== a)| tests its input for equality with \verb|a|.
The \verb|compressSpan| helper function gathers up all elements at the
start of the tail \verb|as| that are equal to the head \verb|a|.  This
prefix is returned as \verb|p|, with the remaining suffix as \verb|s|.
The pair \verb|(1 + length p, a)| is returned to summarize
\verb|a :: p|.  The \verb|mapThrough| combinator then iterates
\verb|compressSpan| through the suffix \verb|s|.

Assuming \verb|A : Set| and an equality test \verb|eqb : A -> A -> bool| on it,
we port this code to our Coq infrastructure in Figure~\ref{fig:rle}.
The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
\verb|compressSpan| has type:
\begin{minted}{coq}
forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> (nat * A) * R.
\end{minted}
\noindent It is invoked by the code for \verb|mapThrough| with \verb|fo : FoldT (Alg (ListF A)) R|.  It then has the
responsibility of extracting from the tail at type \verb|R| (second input)
a result upon which \verb|mapThrough| should recurse (second component
of the output pair).  Then we define an algebra \verb|RleAlg| by
supplying \verb|compressSpan| as the function to map through, to
\verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).  Following the
pattern seen above, we define function \verb|rle| for top-level
recursions using \verb|fold| (we could also define a subsidiary version
\verb|rler|).

>>>>

This first example does not invoke subsidiary recursions, but will
itself be used as a subsidiary recursion in other examples to follow.
Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix. Note that this is the typing for the top-level function.  More
generally, though, given a type \verb|R : Set| and a
fold function for \verb|R| (i.e., of type
\verb|FoldT (Alg (ListF A)) R|), we want to map an input list of
type \verb|R| to a pair of type \verb|list A * R|.  The first
component of this pair is going to be built up from scratch, and so
cannot have type \verb|R|; we cannot statically ensure that outer
recursions on it are legal.  Its second component will always be a
subdatum of the input list, and can thus have the type \verb|R|,
enabling outer recursive calls.  So we want:
\begin{minted}{coq}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{minted}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{minted}{coq}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{minted}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|ListF A R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{minted}{coq}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{minted}
\noindent The idea is that the recursion will return
\verb|SpanNoMatch| to signal that it is in the one tricky case where
\verb|p| does not match the first element.  Otherwise, it will be able
to return, via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This suffix is available to the algebra
in the subdata structure of type \verb|ListF A R|.

@@

Figure~\ref{fig:spanalg} shows the algebra \verb|SpanAlg|, whose type
is \verb|Alg (ListF A) SpanF|.  So we are defining an algebra
(\verb|Alg|) for the \verb|ListF A| functor, with carrier \verb|SpanF|
of the required type \verb|Set -> Set|.  We use \verb|rollAlg| to
create an algebra from something whose type is an application of
\verb|AlgF|.  This takes in all the components of the recursion
universe: the abstract type \verb|R|, the fold function (\verb|fo|)
for any subsidiary recursions (not used here), a function we choose to
name \verb|span| for making recursive calls, and finally
\verb|xs : ListF A R|.  The algebra does case analysis on this last
argument.  In the cases where it is empty or where its head
(\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
This signals to the caller that we really wished to return
\verb|([],xs)|, but could not because we do not have \verb|xs| at type
\verb|R|.  If the head does satisfy \verb|p|, then we recurse on the
tail (\verb|tl : R|) by calling the provided
\verb|span : R -> SpanF R|.  If \verb|span tl| returns
\verb|SpanNoMatch|, that means that we should make \verb|tl| the
suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
have \verb|tl : R| here, so we can do this.  For either possible
return value of \verb|span tl|, we add the current head \verb|hd| to
the front of the prefix.

@@

\subsection{The \texttt{wordsBy} function (\texttt{WordsBy.v})}
We now consider how to write the \verb|wordsBy| function from
Section~\ref{sec:intro}, by using \verb|breakr| subsidiarily.  The code
is in Figure~\ref{fig:wordsby}, assuming a type \verb|A : Set|.  The
setup is similar to that for \verb|span|.  We first define an algebra
\verb|WordsByAlg|, parametrized by a predicate \verb|p|, of type
\verb|Alg (ListF A) (Const (list (list A)))|.  This says that
\verb|WordsByAlg p| is an algebra (\verb|Alg|) for the \verb|ListF A|
functor, with carrier \verb|Const (list (list A))|.  \verb|Const| is
a combinator for creating the object part of constant functors;
\verb|FunConst| creates the morphism part (i.e., the \verb|fmap|
function).  We use \verb|Const| where the return type of the algebra
will not depend on its abstract type \verb|R|.  Since we are
constructing a list of lists from scratch, it will not be legal to
recurse on the list itself, or its (list) elements.  So we just use
the \verb|list| type of Coq's standard library.

The code for \verb|WordsByAlg| is essentially the same as what we saw
in Section~\ref{sec:intro}.  Recall that this function drop elements
which satisfy \verb|p|, and returns a list of the sublists between
maximal sequences of such elements. This algebra pattern matches on
\verb|xs : ListF A R|. In the \verb|Cons| case, if the head
(\verb|hd|) satisfies the predicate, then we drop it and recurse.
Note that this recursive call is legal because \verb|tl| and
\verb|wordsBy| have type \verb|R| and \verb|R -> list (list A)|,
respectively. Otherwise , we use \verb|breakr| to obtain the maximal
prefix \verb|w| of \verb|tl| that does not satisfy \verb|p|, and the
remaining suffix \verb|z|.

Here we see the benefit of our approach.  From Figure~\ref{fig:span},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
comes from the type \verb|FoldT (ListF A) Alg R| of
\verb|fo|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  Thus, it is
legal to recursively apply \verb|wordsBy : R -> list (list A)| to
\verb|z|. The figure also shows the code for the subsidiary
recursion \verb|wordsByr|.

\begin{figure}
\begin{minted}{coq}
Definition WordsByAlg (p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs =>
      match xs with
        Nil => []
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
\end{minted}
\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
  fold (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

This example shows how to write a combinator that factors out a
subsidiary recursion.  The Haskell library \verb|Data.List.Extra| defines
a function \verb|repeatedly|, in essentially the same way as \verb|mapThrough| in
Figure~\ref{fig:mapthroughhs}, although we attempt a more informative
name.  This function behaves like the standard \verb|map| function on lists, except
that the function \verb|f| that we are mapping (or ``mapping
through'') takes in not just the current element \verb|a|, but also
the tail \verb|as|.  It then returns the value \verb|b| to include in
the output list, and whatever other list it wishes, upon which
\verb|mapThrough| will then recurse.

To write this combinator using our infrastructure for subsidiary
recursion, we need to supply the mapped function with the fold
function for \verb|mapThrough|'s recursion.  This is so that the
mapped function can initiate a subsidiary recursion, returning a value
in the abstract type \verb|R| of \verb|mapThrough|'s recursion.
The type we will use for mapped functions is:
\begin{minted}{coq}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{minted}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input.

Given this definition, the Coq definition of \verb|mapThrough| is
presented in Figure~\ref{fig:mapthrough}.  \verb|MapThroughAlg| is
similar to the Haskell code above, though when we call \verb|f|, we
must supply the abstract type \verb|R| and fold function \verb|fo|.
From the definition of \verb|mappedT|, we have that \verb|b : B| and
\verb|c : R|, so we may indeed invoke \verb|mapThrough : R -> list B|
on \verb|c|.  Note that as we are building up a new list from scratch
(rather than just extracting some tail of the input list), we just
return \verb|list B|; we cannot perform further subsidiary recursion
on the output.

\begin{figure}
\begin{minted}{coq}
Definition MapThroughAlg{B : Set}(f:mappedT A B)
  : Alg (ListF A) (Const (list B)) :=
  rollAlg (fun R fo mapThrough xs =>
    match xs with
      Nil => []
    | Cons hd tl =>
      let (b,c) := f R fo hd tl in
        b :: mapThrough c
    end).
Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                      {B : Set}(f:mappedT A B) : R -> list B.
Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B.
\end{minted}
\caption{The algebra \texttt{MapThroughAlg} defining the functions
  \texttt{mapThrough} and \texttt{mapThroughr}. We omit the code for
  those two functions as it follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby})}
\label{fig:mapthrough}
\end{figure}

@@

Finally, we have an example using our \verb|mapThrough| combinator
together with a subsidiary recursion, to implement a \emph{run-length
encoding} of lists.  This is a basic data-compression algorithm where maximal
sequences of $n$ occurrences of element $e$ are summarized by the pair
$(n,e)$~\cite{datacomp}.  The Haskell code for this function is shown in Figure~\ref{fig:rlehs}.
Recall that \verb|(== a)| tests its input for equality with \verb|a|.
The \verb|compressSpan| helper function gathers up all elements at the
start of the tail \verb|as| that are equal to the head \verb|a|.  This
prefix is returned as \verb|p|, with the remaining suffix as \verb|s|.
The pair \verb|(1 + length p, a)| is returned to summarize
\verb|a :: p|.  The \verb|mapThrough| combinator then iterates
\verb|compressSpan| over the suffix \verb|s|.

Assuming \verb|A : Set| and an equality test \verb|eqb : A -> A -> bool| on it,
we port this code to our Coq infrastructure in Figure~\ref{fig:rle}.
The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
we see that \verb|compressSpan| has the type:
\begin{minted}{coq}
forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> (nat * A) * R.
\end{minted}
\noindent It is invoked by the code for \verb|mapThrough| with
\verb|fo : FoldT (Alg (ListF A)) R|.  Note that \verb|compressSpan|
has the responsibility of extracting from the tail at type \verb|R|
(second input) a result upon which \verb|mapThrough| should recurse
(second component of the output pair). Then we can define an algebra
\verb|RleAlg| by supplying \verb|compressSpan| as the function to map
through, to \verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).
Following the pattern seen above, we can write the top-level function
\verb|rle| using \verb|fold| (we could also define a subsidiary
version \verb|rler|).
